Efficient Parallelization of $\ell_1$-Minimization for Face Recognition \thanks{Corresponding author: . This work was partially supported by ARO MURI W911NF-06-1-0076.}

### 2011-02-02 paper/main.tex
fixed merger (i think...)

%The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
%\begin{equation}
%F(x) = f(x) + \lambda g(x).
%\end{equation}

%Because the optimal lagrange multiplier values are not known, ALM iterates between estimating the solution and the lagrange multipier values.

%In order to solve the recognition stage, we consider the $l_1$ problem:
%\begin{equation}
%\min_{\x, \e} \|\x\|_1 \quad \subj \quad \bb = A \x + \e.
%\end{equation}
%which can be formulated as
%\begin{equation}
%L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
%\end{equation}

%where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

We define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

We summarize the ALM algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the
largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

% With a reshuffling of the order of operations and a change of the termination
% condition, the inner loop takes the form of two matrix-vector multiplications
% and a sequence of vector-vector operations.

@@

\subsection{GPU Hardware Parallelism}

@@

\subsection{Single Instance GPU Implementations}

The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU portion
does all of the actual computation.

@@

To construct the benchmark for synthetic data, we first created an $m \times n$
matrix $A$ with entries $\pm 1$ each with a $50\%$ chance.  Then we created a
sparse vector $\x_0$ with $k\%$ nonzero random entries.  We also created a
sparse vector $y_k$ with the same number of non-zero entries as $\x$.  Let
$\y_0 = A \x_0 + \y_k$.  We timed the time taken for the ALM algorithm using
$\y_0$, $A$, and other inputs.  The other inputs included, maximum number of
inner/outer loop iterations, and inner/outer loop tolerance level.
% give a discusison of how to construct the benchmark: one paragraph for
% protocol on how to construct synthetic data second P for protocol on how to
% construct the face data

To simply the benchmark, we cap the number of inner iterations at 50 and outer
iterations at 5000 and run the algorithm to convergence.  CPU code was run on
an Intel Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a
GTX480 We ran the A matrix with different ratios between m:n from 10:1:5 for
m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.

>>>>

In this section, we describe the general version ALM algorithm and then describe how it can be adapted to solve our problem.

The augmented lagrange multiplier (ALM) method utilizes a popular class of convex techniques, the lagrange multiplier, solving the problem:
\begin{equation}
F(x) = f(x) + \lambda g(x).
\end{equation}
which can be formulated as an augmented Langrange function given by
\begin{equation}
L_u(x,y) = \|x\|_1 + <\y, b - A\x - \e> + \frac{\mu}2 \| b-A\x-\e \|_2^2
\end{equation}

where $\mu > 0$ is a constant that penalizes infeasibility and $y$ is a vector of lagrange multipliers.

For an optimal $y*$ which satisfies the second-order sufficiency conditions, for a sufficiently large $mu$, it can be shown that
\begin{equation}
x* = \arg min L_u(x,y*)
\end{equation}

However, because $y*$ is generally unknown, ALM methods simultaneously estimate the optimal solution and Lagrange multipliers iteratively in the following procedure.

We also need to define the following soft-thresholding operator for a
scalar $x$ and a scalar $\alpha \geq 0$:
\begin{equation}
\textup{shrink}(x,\alpha) = \textup{sign}(x)\cdot \max \{|x| - \alpha, 0\},
\end{equation}
In practice, this operator will be applied elementwise to a vector $\x$ with a single $\lambda$,
and will be notated as $\textup{shrink}(\x,\alpha)$.

We summarize the entire ALM
algorithm as Algorithm~\ref{alg:alm}, where $\gamma$ denotes the largest eigenvalue of the matrix $A^TA$. For the choice of parameter $\mu$, we take the same strategy as
in \cite{YangJ2009-pp} and set $\mu_0 = 2m / \|\bb\|_1$. We set $\rho=1.5$.
\begin{algorithm}[h]
\caption{\bf (Augmented Lagrange Multiplier Method Used in Alignment Inner Loop)}
\begin{algorithmic}[1]
\STATE {\bf Input:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$,
$\x_1 = \mathbf{0}$, $\blamda_1 = \mathbf{0}$.
\WHILE{not converged ($k = 1,2,\ldots$)}
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\e_{l+1} \leftarrow \textup{shrink}\left(\bb - A\x_l + \frac{\blamda_k}{\mu_k}, \frac{1}{\mu_k}\right)$;
\STATE $\x_{l+1} \leftarrow (A^\dagger)^T \left(\bb - \e_{l+1} + \frac{\blamda_k}{\mu_k} \right) $;
\ENDWHILE
\STATE $\blamda_{k+1} \leftarrow \blamda_k + \mu (\bb - A\x_{k+1} - \e_{k+1})$;
\STATE $\mu_{k+1} \leftarrow \rho\mu_k$;
\ENDWHILE \STATE
{\bf Output:} $\x^* \leftarrow \x_k, \e^* \leftarrow \e_k$.
\end{algorithmic}
\label{alg:alm}
\end{algorithm}

%Show Pseudocode for one variant
%\subsection{Computational Complexity}
The Matrix-vector computions are O($mn$), and the vector-vector operations are O($m$), 
in terms of both computation cost and memory-bandwidth.  As A'A can be precomputed once, assuming the iterative solver always takes a constant number of iterations, the PALM algorithm has a computational complexity of O($mn+n^2$), where n $<$ m.

% Complexity of entire algorithm

\subsection{Adapations}

In this section, we describe how to adapt the general ALM algorithm for the alignment and recognition problem.  Refer to book...

In order to solve the alignment stage before recognition, we consider the $l_1$ problem: 
\begin{equation} 
\min_{\x, \e} \|\e\|_1 \quad \subj \quad \bb = A \x + \e.
\end{equation}
which is solved in a similar fashion.

\section{CPU Implementation}
% With a reshuffling of the order of operations and a change of the termination
% condition, the inner loop takes the form of two matrix-vector multiplications
% and a sequence of vector-vector operations.

@@

\subsection{GPU Hardware Parallelism}
High-level description of GPU parallelism, i.e.

The ALM algorithm requires many matrix-vector operations, this gives manycore processors such as GPUs a computational advantage for large problem sizes.  
GPUs are comprised of several streaming multiprocessors (SMP) (16 for the Nvidia Fermi), which each contain 32 floating-point units.  Each SMP contains its own L1-cache and shared memory and all SMPs share a common L2-cache.  The GPU executes programs as SIMT (Single Instruction Multiple Threads) in groups of 32 threads called warps.  These warps are then grouped into higher levels called thread blocks and grids.
%The GPU contains an inverted memory hierarchy so that the L1 cache stays close to the cores, providing the increase in performance.  It essentially turns the GPU from being memory-limited to computational-limited.
GPU memory also differs from CPU memory in that it requires data transfers over PCI-Express, which have limited bandwidth.  The cost of these transfers are amortized by doing large amounts of computation at once, requiring code to be run completely on the GPU.  The GPU can be analogous to a multi-core CPU processor, except with more sophisticated vector processing hardware.  Each thread block is a virtualized multiprocessor, which each thread as a virtualized scalar processor.  All of these differences allow GPUs to perform intensive calculations such as matrix-vector operations at much faster rates than CPUs.

%Each SMP is analogous to a single CPU except that it can run multiple threads simultaneously.

%You should give a brief (1-2 paragraphs) overview of the GPU architecture -- most people have heard of GPUs and what they can do by now, so it isn't worth spending a whole lot of space in the paper describing it.

%You should define the levels of the Cuda thread and memory hierarchies: i.e. warps, thread blocks, grids and register file, shared memory, global memory. Also mention that the GPU memories are all distinct from the CPU memories, and data transfers go over PCI-Express -- but the cost of those transfers are amortized by doing lots of work (i.e. solving an $\ell_1$ minimization after transferring the matrix once). It is worth spending a little space describing how the GPU is a multi-core processor (just like the CPUs), but with much more sophisticated vector processing hardware. That is: on a Fermi there are 15 cores, and each one has two 16-wide vector floating-point units.

%The GPU-algorithm is the same as the CPU algorithm: primary- and dual- ALM methods. The implementations differ in that you have to parallelize the GPU code at an additional level, and you can't use the GPU's caches to communicate between cores (only within a core). On the CPU, you only need to parallelize for the 4 or 8 cores -- and perhaps add SSE instructions when possible (usually this is inside of the BLAS library) -- any communication between cores happens implicitly via cache coherence. For the GPU, you need to parallelize for the 15 cores, but also for the 32-way SIMD parallelism of Warps. Any intra-core communication happens via the cache (aka scratchpad, or __shared__ memory) and within-block barriers (i.e. __syncthreads()), and any inter-core communication happens via the GPU's DRAM and the implicit barriers between grid launches.

%cuda cores, SIMD width, scheduling restrictions, bandwidth
\subsection{Solving a Single $\ell_1$ minimization}

@@

\subsection{Single Instance GPU Implementations}
The GPU ALM algorithm is written in CUDA and utilizes the CUBLAS libraries.
CUBLAS are the BLAS libraries consisting of all single precision BLAS calls
which are provided by Nvidia and are optimized for GPU performance. In
addition, we coded kernel (GPU) methods for vector-vector operations to replace
operations that would have taken multiple BLAS calls. Because data transfer from the RAM to GPU suffers bandwidth limitations, we designed our implementation to transfer the A matrix and measurement vector at once.  The algorithm is similar
to the CPU algorithm except that we had to parallelize the code at an
additional level. The CPU portion of the code serves as the conductor, making
calls to GPU functions and directing the flow of the code while the GPU keeps all of the data and runs the computation.

@@

To construct the benchmark for synthetic data, we first created an $m \times n$
matrix $A$ with entries $\pm 1$ each with a $50\%$ chance.  Then we created a
sparse vector $\x_0$ with $10\%$ nonzero random entries.  We also created a
sparse vector $y_k$ with the same number of non-zero entries as $\x$.  Let
$\y_0 = A \x_0 + \y_k$.  We timed the time taken for the ALM algorithm using
$\y_0$, $A$, and other inputs.  The other inputs included, maximum number of
inner/outer loop iterations, and inner/outer loop tolerance level.
% give a discusison of how to construct the benchmark: one paragraph for
% protocol on how to construct synthetic data second P for protocol on how to
% construct the face data

\subsection{PALM $\ell_1$ minimization benchmark on Synthetic Data}

In this section, we benchmark the performance of BLAS MKL and GPU CUDA
implementations of the Primal ALM algorithm on synthetic data to find at which
point the GPU implementation runs faster than CPU implementation.  To simply
the benchmark, we cap the number of inner iterations at 50 and outer iterations
at 5000 and run the algorithm to convergence.  CPU code was run on an Intel
Dual Socket Quad Core with 8 gigs of DDR3 RAM.  GPU code was run on a GTX480 We
ran the A matrix with different ratios between m:n from 10:1:5 for m:2000:25:8000 with an x vector of 10 percent sparsity.  Tolerance levels were
empirically seen to allow best performance with the inner loop tolerance set at
one order of magnitude lower than the outer loop tolerance.
